---
title: "03a_Phylogenetics"
author: "sophiaaredas"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # Always relevant to the document directory 
                      # send any figure output to this folder 
                      fig.path = "../figures/03_Phylogenetic_Tree/",
                      warning = FALSE) 
```

---
# Goals of this file 

The goal is to create a phylogenetic tree! 

1. Load in preprocessed phyloseq object.  
2. Create ASV fasta file from the phyloseq object. 
3. Align the 16S sequences from fasta file with MAFFT. 
4. Create a tree with FastTree2.  

# Set the seed 
```{r set-seed}
set.seed(0909199)
```


## Load Packages & Phyloseq Object
```{r load-environment}
# phytools, ggtree, RColorBrewer
pacman::p_load(phytools, ggtree, RColorBrewer, install = FALSE)

# Load physeq 
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# Create Fasta File of ASV and their Sequences

This fasta file will be used to create our alignment in MAFFT

```{r create-preprocessed-fasta}
# pull out ASV seqs and ASV names 
asv_seq_df <- 
  raw_preprocessed_physeq@tax_table %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseq)

#View(asv_seq_df)

# Add the > to make fasta header 
asv_seq_df$ASV <- paste0(">",asv_seq_df$ASV)
#View(asv_seq_df)

# Create the fasta object 
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseq))
head(asv_seq_fasta)

# Write to a file 
write(asv_seq_fasta, 
      file = "data/03_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Align the 16S sequences from fasta file with MAFFT

- `engine.opts = '-l'`: give us original terminal base environment.

```{r run-mafft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run mafft 
# First provide the path to mafft 
export PATH=/programs/mafft/bin:$PATH
# change directories to provide the fasta file we made above 
cd data/03_Phylogenetic_Tree/
pwd

# Set a seed  - using same seed as before for consistency 
RANDOM=1234

# Run Mafft 
# To test in the shell directly from Rmd 
# mac: command + option + enter 
# Windows: control + alt + enter 
# For now, use default options, note the version 
# MAFFT automatically knows that it's a nucleotide alignment 
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta

# Change back to the project directory 
cd ../../
pwd
```



# FastTree2 
```{r run-fasttree, engine='bash', engine.opts="-l"}
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# cd into the alignment file folder 
cd data/03_Phylogenetic_Tree/
pwd

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

# Change back to the main directory 
cd ../../
pwd 
echo "The working directory is $PWD"
```



# Session Information
```{r session-info}
devtools::session_info()
```