---
title: "03b_Phylogenetics"
author: "sophiaaredas"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

# Goals 

1. Load the fastree unrooted tree.  
2. Add tree to phyloseq object.  
3. Visualize and inspect tree with ggtree. 
4. Prune ASVs, if needed.  
5. Root our tree. 
6. Combine new tree with a phyloseq object. 
7. Save 2 phyloseq objects: 1. Unrooted tree phyloseq object, 2. Rooted tree phyloseq object. 

# Before you start

## Set my seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```

## Load Packages 
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools,
               install = FALSE)
```

## Load Data files 
```{r load-data}
# Preprocessed phyloseq object 
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Load in the tree! 
unrooted_tree <- read.tree("data/03_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)
```


# Merge Phyloseq 
```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))

# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)
unrooted_physeq
```

# Plot Tree with `ggtree`
```{r plot-tree-unrooted}
# Make a basic tree
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(size = 0.1, mapping = aes(color = Kingdom)) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom"); kingdom_tree

kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -2, vjust = 1, size = 0.5)
kingdom_node_tree

phylum_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by phylum 
  geom_tippoint(size = 0.1, mapping = aes(color = Phylum)) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "none"); phylum_tree

phylum_node_tree <- 
  phylum_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -2, vjust = 1, size = 0.5)
phylum_node_tree
```
# Evaluate the crazy tree 
This part gets a bit manual
We will be pruning ASVs whose phylum is NA
```{r prune-NA-taxa}
# Function from Joey McMurdie: https://github.com/joey711/phyloseq/issues/652
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

# get rid of NA phylum
no_asvs<-
  unrooted_physeq %>% 
  tax_table() %>% 
  data.frame() %>% 
  dplyr::filter(is.na(Phylum)) %>% 
  pull(ASV)

no_asvs

# Function from Joey McMurdie: https://github.com/joey711/phyloseq/issues/652
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

# Let's use the pop_taxa function :) 
# Recreate a phyloseq object without ASV_456
unrooted_physeq_rmNA <- 
  unrooted_physeq %>%
  pop_taxa(., no_asvs)

#intiution check
ntaxa(unrooted_physeq) - ntaxa(unrooted_physeq_rmNA)

```


# Midroot Tree
```{r midroot-tree}
# Is the tree rooted?
new_unrooted_tree <-phy_tree(unrooted_physeq_rmNA)
is.rooted(new_unrooted_tree)

# Let's midpoint root the tree
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)

# Is the new tree rooted?
is.rooted(midpoint_rooted_tree)

# Assign to a new phyloseq object: merging subsetted phyloseq with the new rooted tree

# Merge tree 
midroot_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, midpoint_rooted_tree)
midroot_physeq

# Quick inspection of tree 
ggtree(midroot_physeq) + 
  geom_tippoint(mapping = aes(color = Phylum))
```


# Save to a new phyloseq object
```{r save-physeq}
# Save both phyloseq objects with our tree object to one .RData file 
save(list = c("unrooted_physeq", "midroot_physeq"),
     file = "data/03_Phylogenetic_Tree/phytree_preprocessed_physeq.RData")
```

# Session Information 
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```