---
title: "02 - PreProcessing"
output: 
  html_document:
    css: "/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis/css/tufte.css"
    toc: yes
    toc_float:
      collapse: no
      smooth_scroll: yes
      toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
setwd("/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis")
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.path = "/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis/figures/02_PreProcessing/") # Send any figures to this folder
```

# Load libraries 
```{r load-libraries}
library(devtools)
#devtools::install_github("joey711/phyloseq")
library(phyloseq)
library(tidyverse)
```

# Goals

Here, we will process the data into a phyloseq object. 

- ASV table 
- Taxonomy Table 
- Track Reads (metadata)

Then, we will remove the following: 

1. Remove Chloroplasts
2. Remove Mitochondria. 
3. Remove samples without "enough" reads. 

Finally, write data file of phyloseq output.


# Load Data 

## ASV Table

```{r load-asv-table}
# First, load asv table
load("/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis/data/01_DADA2/ASV_counts.RData")

# Inspect asv_tab
head(asv_tab)[,1:5]

# Fix names 
sample_names <- colnames(asv_tab)
samples_fixed <- sapply(strsplit(basename(sample_names), "_"), `[`,1) 
head(samples_fixed)

# re-write the ASV count file to fix names 
colnames(asv_tab) <- samples_fixed
str(asv_tab)
```

##  Taxonomy Table
```{r load-tax-table}
tax_df <- read.table("/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis/data/01_DADA2/ASV_taxonomy.tsv", sep = "\t", skip = 1)
head(tax_df)

# fix column names 
colnames(tax_df) <- c("asv_names", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "ASV", "ASVseq")

head(tax_df)

# Taxonomy Table Matrix
tax_mat <- 
  tax_df %>%
  tibble::column_to_rownames(., var = "asv_names") %>%
  as.matrix()
```

##  Track Reads Data
```{r load-track-reads}
load("/local/workdir/zlr6/git_repos/biomi6300_moonmilk_amplicon_analysis/data/01_DADA2/track_read_counts.RData")

# Take a look at the data
head(track_counts_df)
dim(track_counts_df)

# Please in the terminal copy metadata.csv 
# into: SalinityGradient_16S/data 
# from: /workdir/in_class_data/SalinityGradient_16S/metadata.csv

# We don't currently have experimental metadata

# Load in metadata
#rm(metadata_df)
#metadata_df <- read.csv("data/metadata.csv")
#dim(metadata_df)
#colnames(metadata_df)


# Merge metadata_df with track_reads_df 
#head(metadata_df)
#head(track_counts_df)

#metadata_track_reads_df <- 
  #metadata_df %>%
  #left_join(., track_counts_df, by = "names") %>%
  #dplyr::select(-X)

# Intuition check 
#head(metadata_track_reads_df)

# Update row.names to be sample names 
## Before 
#row.names(metadata_track_reads_df)
# Rewrite 
#row.names(metadata_track_reads_df) <- metadata_track_reads_df$names
# Check afterwards that it worked 
#row.names(metadata_track_reads_df)
# intuition check
#head(metadata_track_reads_df)
```

# Handoff to phyloseq
```{r phyloseq-handoff}
# double check it's all good 
dim(asv_tab)
dim(tax_mat)

# Intuition check 
stopifnot(row.names(asv_tab) == row.names(tax_mat))

# Construct the phyloseq object 
raw_physeq <- phyloseq(otu_table(asv_tab, taxa_are_rows = TRUE),
                       sample_data(metadata_track_reads_df), tax_table(tax_mat))
#remember to change sample_Data back once we get metadata
                       
raw_physeq

# Save this raw phyloseq object 
save(raw_physeq, file = "data/02_PreProcessing/raw_physeq.RData")
```

# Session Information 
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```
